---
title: "Data Tools Forum 2015: R and R Studio"
author: "Daina Bouquin and Zac Painter"
output: html_document
---

The following code should be run in R Studio. All corresponding documentation is available on GitHub [here](https://github.com/dbouquin/data_tools_2015).   
This markdown file can also be viewed on R Pubs [here](http://rpubs.com/dbouquin/DataTools_2015).

The following packages may be useful to you but will not be the focus of the demonstrations during the Data Tools Forum:
```{r, warning=FALSE, error=FALSE, message=FALSE}
library(swirl) # This is the tutorial we encourage you to go through prior to the session, specifically the first two modules
library(haven)
library(xlsx)
```

These packages will be used and discussed during the Wrangling portion of the presentation:
```{r, warning=FALSE, error=FALSE, message=FALSE}
library(dplyr) # We will use this package during demonstrations
library(tidyr) # more data restructuring
library(stringr) # for regular expressions
```

These packages will be used and discussed during the Analysis through data visualization portion of the presentation:
```{r, warning=FALSE, error=FALSE, message=FALSE}
library(ggplot2) # We will use this package during demonstrations
library(caret) # for classification and regression training
library(zoo) # for time series data
```

<h4>Data Wrangling (AKA "munging")</h4>   
We're going to be using the 'diamonds' dataset, which can be accessed by installing and loading `ggplot2`
   
```{r}
# Check out the structure of the dataset
# str() function
str(diamonds) 

# head() for first few records
head(diamonds)
```
   
Diamonds munging:
```{r}
# Subsetting out first 10 rows
diamonds[1:10,]          

# Asigning the first 10 rows to a variable (object) called di_sub
di_sub <- diamonds[1:10,]

# Print the object and check its structure
di_sub
str(di_sub)

# What if we just want to see a few columns of the new dataframe?
# We can use select()
di_sub2 <- select(di_sub, carat, cut, color, price, clarity)
di_sub2

# Try out filtering on the larger dataset for diamonds where the cut is ideal
diamonds_ideal <- filter(diamonds, cut=="Ideal")
head(diamonds_ideal)

# Use mutate to add a variable to the ideal diamonds dataset
diamonds_ideal <- mutate(diamonds_ideal, price_per_carat = price/carat)
head(diamonds_ideal)

# Write one of our new objects to a CSV file in your working directory
file <- write.csv(di_sub,"di_sub.csv")

# Read in the CSV file you just created
di_sub2 <- read.csv("di_sub.csv")

# Subset the dataset you just read in and asign it to a new object
new_diamonds <- di_sub2[,2:10]

# print the new object
new_diamonds
```
    
<h4>Data Analysis through visualization</h4>
```{r}
# Base plotting for comparison
base_scatter <- plot(diamonds$carat, diamonds$cut,
  xlab = "carat", ylab = "cut")
# Print the plot to the panel
base_scatter
# Try another base plot
base_price <- plot(diamonds$carat, diamonds$price)
base_price
# Now we'll check out how base plotting differs from ggplot2 functionality
# Make a box plot
diamonds_box <- ggplot(diamonds, aes(y= carat, x= cut)) + geom_boxplot()
diamonds_box
# How about a violin plot? Just change the geom parameter
diamonds_violin <- ggplot(diamonds, aes(y= carat, x= cut)) + geom_violin ()
diamonds_violin
# Build up the plot with another layer by adding another geopm to the violin plot.
diamonds_vpoint <- diamonds_violin + geom_point()
diamonds_vpoint
```

Heatmap example from [R-Blogger](http://www.r-bloggers.com/simplest-possible-heatmap-with-ggplot2/)
```{r}
library(reshape2)
library(RColorBrewer)

# Data creation and pre-processing
nRow <- 9
nCol <- 16
myData <- matrix(rnorm(nRow * nCol), ncol = nCol)
rownames(myData) <- letters[1:nRow]
colnames(myData) <- LETTERS[1:nCol]

# Replace with numbers that actually have a relationship:
for(ii in 2:ncol(myData)){  myData[, ii] <- myData[, ii-1] + rnorm(nrow(myData))  }
for(ii in 2:nrow(myData)){  myData[ii, ] <- myData[ii-1, ] + rnorm(ncol(myData))  }

# For melt() to work seamlessly, myData has to be a matrix.
longData <- melt(myData)

# Optionally, reorder both the row and column variables in any order
# Here, they are sorted by mean value
longData$Var1 <- factor(longData$Var1, names(sort(with(longData, by(value, Var1, mean)))))
longData$Var2 <- factor(longData$Var2, names(sort(with(longData, by(value, Var2, mean)))))

# Define palette
myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")), space="Lab")

# Make the headmap!
zp1 <- ggplot(longData,
              aes(x = Var2, y = Var1, fill = value))
zp1 <- zp1 + geom_tile()
zp1 <- zp1 + scale_fill_gradientn(colours = myPalette(100))
zp1 <- zp1 + scale_x_discrete(expand = c(0, 0))
zp1 <- zp1 + scale_y_discrete(expand = c(0, 0))
zp1 <- zp1 + coord_equal()
zp1 <- zp1 + theme_bw()
zp1
```
